Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2025-05-21 manuscript/manuscript.tex
Working on discussion

Model performance across 5 neonatal CODs, excluding congenital anomalies (n=2, \textless 1\%) and other (n=5, 1\%) due to small sample sizes is shown in Figure \ref{fig-pccc-cod-neo}. GPT-4 had the highest individual performance for 3 of 5 CODs (0.00-0.00 PCCC), GPT-3.5 for 2 CODs (0.00-0.00 PCCC), and InSilicoVA for 1 COD (0.00-0.00 PCCC).

@@

This section identifies advantages of performant models for assigning CODs. Section \ref{discuss-adv-cod} details the strategic use of models for particular CODs. Section \ref{discuss-adv-hscav} details the resource efficiency of GPT and InSilicoVA models for assisting in physician COD assignment. Section \ref{discuss-adv-nlf} notes the strength of using natural language text in GPT models compared to structured data (e.g. questionnaires) for physician COD assignment.

\subsubsection{Performance for Particular Causes of Death}\label{discuss-adv-cod}

At the population level, overall performances for all causes were similar (0.7 $\leq$ CSMF $\leq$ 0.79), but not close ($\geq$0.8  CSMF), to physicians at 0.74-0.79 CSMF. At the individual level, performance by PCCC varies across CODs, and for most CODs, GPT models (GPT-4 and GPT-3) performed better than InSilicoVA and InterVA-5 (Table \ref{tab-perf-geq80}). However, there were CODs where InSilicoVA performed better. For example, neonatal infections at 0.87 PCCC versus 0.23 for GPT-3.5, and adult tuberculosis at 0.72 versus 0.39 for GPT-3.5. For CODs with high performance ($\geq$0.8 PCCC) (Table \ref{tab-perf-geq80}), the results suggest that GPT models (and InSilicoVA for particular CODs) may assign CODs that are close physicians, when physicians would agree on the COD. Thus, when applying models to assign CODs, it may help to target CODs with a combination of models that performed well or close to physician assignment.

Across age ranges, GPT and InSilicoVA models had performances below 0.8 PCCC, where ages 10-24 years and 7-27 days had the highest performance ($\geq$0.7 PCCC). For adult age ranges, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults (Figure \ref{fig-perf-agerange-adult}). For child and neonatal age ranges, the performance improves as the age increases, suggesting less difficulty in COD assignment when children and neonates are more developed (Figures \ref{fig-perf-agerange-child} and \ref{fig-perf-agerange-neo}). As the models did not perform well for any particular age range, it may not be suggested to apply specific models to target cases by age range. However, the general patterns (increases and decreases of performance in relation to age) are useful in comparing expected patterns in relation to physician assignment.

>>>>

Model performance across 5 neonatal CODs, excluding congenital anomalies (n=2, \textless 1\%) and other (n=5, 1\%) due to small sample sizes is shown in Figure \ref{fig-pccc-cod-neo}. GPT-4 had the highest individual performance for 3 of 5 CODs (0.39-0.71 PCCC), GPT-3.5 for 1 COD (0.57 PCCC), and InSilicoVA for 1 COD (0.86 PCCC). GPT-3.5 had the lowest performance for 3 of 5 CODs (0-0.13 PCCC) and InterVA-5 for 2 CODs (0.01 and 0.48 PCCC). All models had similar performance for stillbirth deaths (0.48-0.57 PCCC), while only GPT-4 had a PCCC \textgreater 0 PCCC. InSilicoVA had improvements over all other models for neonatal infection deaths (+0.18-0.73 PCCC).

@@

This section identifies the advantages of models for assigning CODs. Section \ref{discuss-adv-cod} details the application of models for particular CODs. Section \ref{discuss-adv-hscav} details the resource efficiency of computer models for assisting in physician COD assignment. Section \ref{discuss-adv-nlf} notes the strength of using natural language text in GPT models compared to structured questionnaire data for physician COD assignment.

\subsubsection{Cause and Age Specific Models}\label{discuss-adv-cod}

At the population level, overall performances for all models were similar to physicians (0.74-0.79 CSMF). At the individual level, performance varied across CODs (0-0.99 PCCC). For most CODs, GPT-3.5/GPT-4 performed better than InSilicoVA/InterVA-5 (top PCCC for 15 of 17, 7 of 8, and 4 of 5 adult, child, and neonatal CODs respectively), while, InSilicoVA performed better for particular CODs (road and transport injuries, tuberculosis, pneumonia, and neonatal infections with 0.84, 0.71, 0.78, and 0.86 PCCC respectively). For CODs with high performance ($\geq$0.8 PCCC) (Table \ref{tab-perf-geq80}), the results suggest that GPT models and InSilicoVA may assign CODs that are close physicians, when physicians would agree on the COD. Thus, when applying models to assign CODs, it may help to target CODs with a combination of models that performed well or close to physician assignment.

Across age ranges, all models followed a similar upward trend from 6 months to 14 years of age, and a downward trend from 15-69 years with GPT models having higher performance than InSilicoVA/InterVA-5 models, while more mixed trends were observed from 0 days to 5 months (recall Figure \ref{fig-pccc-age}). For adult age ranges, performance generally decreased as age increased, which suggested that models had difficult assigning CODs for older than younger adults with some improvements after the age of 59. For child and neonatal age ranges, the performance improves drastically as the age increases after 5 months, suggesting less difficulty in COD assignment when children and neonates are more developed. As the models did not perform particularly well (\textgreater 0.8 PCCC) for any particular age range, it may not be suggested to apply specific models to target cases by five-year age range. However, the patterns of increases and decreases of performance in relation to age are useful in comparing expected patterns in well-studied medical literature and knowledge.
